A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-23 tex/body.tex
more edits

To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is less than $u$ based on a term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.

@@

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

For each goal in each theory, Sledgehammer was invoked to select up to
256~lemmas, which were monomorphized and translated to SMT-LIB~2
along with the goal \cite{blanchette-et-al-2013-smt}. The resulting problem was
given to CVC4, running for up to 60 seconds on the StarExec cluster
\cite{stump-et-al-2014-starexec}.
%TODO: Keep or not? Problems not involving (co)datatypes were filtered out.
Due to the lack of machinery for reconstructing inferences about (co)datatypes
in Isabelle, the solver is trusted as an oracle in these experiments.

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning. Z3 4.3.2 is included for comparison.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

>>>>

To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is smaller than $u$ with respect to a term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.

@@

\kern-1pt

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We also

%%% TYPESETTING: Terrible hack
\pagebreak
\noindent
added
two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before conducting the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

For each goal in each theory, Sledgehammer was invoked to select up to
256~lemmas, which were monomorphized and translated to SMT-LIB~2
along with the goal \cite{blanchette-et-al-2013-smt}. The resulting problem was
given to the development version of CVC4 and to Z3 4.3.2, each running for up to 60
seconds on the StarExec cluster \cite{stump-et-al-2014-starexec}.
%TODO: Keep or not? Problems not involving (co)datatypes were filtered out.
Due to the lack of machinery for reconstructing inferences about (co)datatypes
in Isabelle, the solvers are trusted as oracles in these experiments. Z3
is included for comparison.

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}
